Method

### 2025-11-25 GroupPolarizationStatmod/paper/method.qmd
magic expanding methods section

---
title: Method
editor: 
  markdown: 
    wrap: 80
author: Matt Turner · [https://mat.phd](https://mat.phd)
date: today
toc: true
bibliography: this.bib
---

@@

Throughout this paper we model a participant $i$'s latent opinion at time $t$ as
a normally-distributed random variable,
$$ 
\omega_{i,t} \sim
\mathcal{N}(\mu_t, \sigma_t),
$$ {#eq-random-variable-def}
with mean $\mu_t$ and standard deviation $\sigma_t$. 
Group polarization in terms of latent opinions is then defined as any pair of pre- and
post-discussion opinion distributions where the mean increases in magnitude over
time—formally, $| \mupost | > | \mupre |$. We need to take magnitudes since radicalization may be towards increasingly negative or positive opinion values.

For group polarization experiments, only two times are relevant:
*pre-discussion*, $t=\mathrm{pre}$, and *post-discussion*, $t=\mathrm{post}$. Since participant groups are assumed to be likeminded, we could also assume they come to agree more over time as well as become more extreme in our group polarization model, or in terms of the latent opinion variance, $\sigmapre > \sigmapost$—but only one study in our ten actually measured opinion variance to check this [@Schkade2010].

@@

## Two Ways to Model Measurement that Extend the Latent Opinion Model

We simulate measurement in two different ways—i.e., with two different models—for the two parts of our analysis. The first
is the *expectation model*: an
efficient way to *simulate measurement of the observed average* opinion of a group, using the *mathematical expectation value* of the distribution of measured opinions as a proxy. In the second
model, the *stochastic model*, we simulate individual participant responses by first drawing an opinion for each participant from the latent opinion distribution (@eq-. The opinions are binned to the appropriate ordinal value.

@@

# *Identification* of Invalid Replications

In the first stage of our analysis, the *Identification* stage, we check if we can *invalidate* a replication of group polarization achieved in some experiment/experimental condition, which we label $e$. Our invalidation strategy is to find a mere agreement null-polarization model, call it $\nu$, that predicts the exact observations reported in each experimental replication in our corpus. If one is found, then we have found a model consistent with the *negation* of the group polarization hypothesis that explains the data as well as the group polarization hypothesis itself. In other words, the hypothesis is not proved or disproved given the data, but *undecidable*: the data cannot *decide* the hypothesis one way or another because of potential measurement artifacts. Because the experiment failed to produce a valid replication, the results are effectively stripped of evidentiary value and warrant exclusion from the aggregate evidence base.

@@

So, to review, this subsection introduces how the *Identification* stage of the analysis works. In this stage we identify for which experiments, $e$, does a mere agreement null model look like group polarization under $e$'s measurement procedure.

## The Expectation Model of Opinion Measurement

@@

To simulate the measurement of the *average* observed opinion at time $t$, $\bar{o}_t$, as the expected value of the probability density function of observed ordinal opinions. In other words, we simulate $\bar{o}_t$ by calculating the weighted sum of products between the probability of observing each bin value and the bin value itself:
$$
\bar{o}_t(\nu, e) = \sum_{i=1}^B b_i f_{it}(\nu, e).
$${#eq-sim-ave-observed}

Group polarization is the difference between the post-deliberation mean observed
opinion and the pre-deliberation one. We will calculate the magnitude of spurious group polarization under a mere agreement null model represented by its collected distribution parameters, $\nu = (\mu, \sigmapre, \sigmapost)$, under experimental design, $e$: 
$$
g(\nu, e) = |\bar{o}_\mathrm{post}(\nu, e)| - |\bar{o}_\mathrm{pre}(\nu, e)|
$${#eq-spurious-gp}

@@

## Example: the "Americans" condition in Moscovici and Zavalloni (1969)

@@

Since the simulated average opinions match the observed ones—note that as a consequence, the spurious polarization magnitude also matches the observed one.
In other words, we *identified* a null model that explains the data. 
If both a hypothesis and its null explain data equally well, then the hypothesis is *undecidable*

In this example we started with

## Root-finding Algorithm to Find Null-Polarization Models with Artifacts

# Evaluation of Experimental Designs

- The evaluation stage complements our *Identify* stage by filling in more details about the experimental design.

@@

First, how frequently do the $\nu$ that we found generate simulated data
  that fool a rigorous statistical design that properly accounts for the measurement process? This serves two roles: 
  1. Confirm that the $\nu$ we found could create simulated datasets that appear polarized under statistical analysis, which is different from using the expected value of an exact latent opinion distribution, integrated with the bin values to simulate the average opinion.
  2. Evaluate the underlying experimental design independent of the failure to account for how ordinal measurements inevitably clip extreme latent opinions.
  3. We do this by simulating experiment data using $\nu$ we identified in the previous stage. We simulate 1000 experiments indexed by $i$ for each experiment $e$.

@@

Each trial $i$ of $N_e$ simulated pre- and post-deliberation observations is then inspected to see
if spurious group polarization occurred in the simulation. Spurious group
polarization is determined by first inferring the latent parameters of each
simulated experimental trial by fitting an ordered probit model~\cite{Liddell2018}. The procedure infers the latent distribution parameters that we in fact specified with $\nu$: $\mu_\mathrm{pre}$, $\sigma_\mathrm{pre}$, $\mu_\mathrm{post}$, and $\sigma_\mathrm{post}$.

The inferred parameters are then used to calculate the effect size for that
trial, Cohen's $d$, defined as~\cite[p. 331]{Liddell2018}
$$
  d = \frac{\mupost - \mupre}{\sqrt{\frac{1}{2}(\sigmapost^2 + \sigmapre^2)}}.
$$ {#eq-cohens}
It is implied that these values are the inferred values, not the values from $\nu$, when used in the context of calculating Cohen's $d$. Note $d=0$ always for the true $\nu$.

@@

To calculate the family-wise error rate
across all experimental conditions of a study, $s$, 
\begin{equation}
  \alpha_s = \frac{1}{|s|} \sum_{e \in s} \alpha_e,
\end{equation}
\noindent
where $e \in s$ indicates sum over all experimental conditions $e$ presented in the article associated with study $s$.
The mean across all results identified as plausibly spurious is 
\begin{equation}
  \alpha_\mathrm{all} = \frac{1}{|E|} \sum_{e \in E} \alpha_e.
\end{equation}

@@

We finally may calculate the false discovery rate by plugging this $\alpha$
into Equation~\ref{eq:fdr} and assuming empirically-motivated values for 
base rate, $b$, and statistical power, $W$, as well as more pessimistic and
optimistic settings for base rate and power.

Using the notation outlined above, the false detection rate is
\begin{equation}
  \fdr = \frac{(1 - b)\alpha}{(1 - b)\alpha + bW}.
  \label{eq:fdr}
\end{equation}

## *Specify* Significance Threshold for Low False Positive Rates

>>>>

---
title: Method
editor: 
  markdown: 
    wrap: 80
author: Matt Turner · [https://mat.phd](https://mat.phd)
date: today
toc: true
bibliography: this.bib
format:
  html:
    grid:
      body-width: 880px  
---

@@

Throughout this paper we model a participant $i$'s latent opinion at time $t$ as
a normally-distributed random variable, $$ 
\omega_{i,t} \sim
\mathcal{N}(\mu_t, \sigma_t),
$$ {#eq-random-variable-def} with mean $\mu_t$ and standard deviation
$\sigma_t$. Group polarization in terms of latent opinions is then defined as
any pair of pre- and post-discussion opinion distributions where the mean
increases in magnitude over time—formally,
$| \mu_\mathrm{post}| > | \mu_\mathrm{pre}|$. We need to take magnitudes since
radicalization may be towards increasingly negative or positive opinion values.

For group polarization experiments, only two times are relevant:
*pre-discussion*, $t=\mathrm{pre}$, and *post-discussion*, $t=\mathrm{post}$.
Since participant groups are assumed to be likeminded, we could also assume they
come to agree more over time as well as become more extreme in our group
polarization model, or in terms of the latent opinion variance,
$\sigma_\mathrm{pre}> \sigma_\mathrm{post}$—but only one study in our ten
actually measured opinion variance to check this [@Schkade2010].

@@

# Two Models of Group Polarization Measurement

We simulate measurement in two different ways—i.e., with two different
models—for different parts of our analysis. The first is the *expectation
model*: an efficient way to *simulate measurement of the observed average*
opinion of a group, using the *mathematical expectation value* of the
distribution of measured opinions as a proxy. In the second model, the
*stochastic model*, we simulate individual participant responses by first
drawing an opinion for each participant from the latent opinion distribution
(@eq-random-variable-def). The opinions are binned to the appropriate ordinal
value.

@@

# Part 1: *Identify* Invalid Replications

In the first stage of our analysis, we try to *identify* a null polarization
models, $\nu$, that invalidate published detections of group polarization if the
measurement procedure induces *spurious group polarization* for that null
polarization model. Null polarization can take many forms—all that is required
is that the latent mean opinion stays constant before and after discussion. Here
we use a *mere agreement* null polarization model, where we simulate
*consensus*, the tendency for sufficiently similar opinions to become more
similar—but we leave out polarization. In mere agreement, moderate and extreme
opinions approach change by equal magnitude towards each other. We therefore
represent consensus as a decrease in the standard deviation of the latent
opinion distribution from pre- to post-discussion. We simulate the measurement
of the group's average opinion before and after discussion and measure the
spurious group polarization, if there is any. We use this procedure in a search
algorithm we developed to identify which combination of constant latent mean and
pre- and post-discussion opinion variances generate spurious group polarization,
if there are any.

Our invalidation strategy, then, is to find a mere agreement null polarization
model, $\nu$, that predicts the exact observations reported in each experimental
replication in our corpus. If one is found, then we have found a model
consistent with the *negation* of the group polarization hypothesis that
explains the data as well as the group polarization hypothesis itself. In other
words, the hypothesis is not proved or disproved given the data, but
*undecidable*: the data cannot *decide* the hypothesis one way or another
because of potential measurement artifacts. Because the experiment failed to
produce a valid replication, the results are effectively stripped of evidentiary
value and warrant exclusion from the aggregate evidence base. We apply our
analysis to every experimental condition across all ten journal articles in our
corpus, which we label $e$.

@@

So, to review, this subsection introduces how we identify which experimental
data can be explained by a mere agreement null model, where the ordinal survey
measurement procedure induces a clipping artifact that hides changes in extreme
opinions that fall outside the measurement scale.

## The Expectation Model of Spurious Group Polarization

@@

To simulate the measurement of the *average* observed opinion at time $t$,
$\bar{o}_t$, as the expected value of the probability density function of
observed ordinal opinions. In other words, we simulate $\bar{o}_t$ by
calculating the weighted sum of products between the probability of observing
each bin value and the bin value itself: $$
\bar{o}_t(\nu, e) = \sum_{i=1}^B b_i f_{it}(\nu, e).
$$ {#eq-sim-ave-observed}

Group polarization is the difference between the post-deliberation mean observed
opinion and the pre-deliberation one. We will calculate the magnitude of
spurious group polarization under a mere agreement null model represented by its
collected distribution parameters,
$\nu = (\mu, \sigma_\mathrm{pre}, \sigma_\mathrm{post})$, under experimental
design, $e$: $$
g(\nu, e) = |\bar{o}_\mathrm{post}(\nu, e)| - |\bar{o}_\mathrm{pre}(\nu, e)|
$$ {#eq-spurious-gp}

@@

## Example: Invalidating the "Americans" condition in Moscovici and Zavalloni (1969)

@@

Since the simulated average opinions match the observed ones—note that as a
consequence, the spurious polarization magnitude also matches the observed one.
In other words, we *identified* a null model that explains the data. If both a
hypothesis and its null explain data equally well, then the hypothesis is
*undecidable* under the available data. This invalidates this experiment's claim
to be a replication of support for the group polarization hypothesis. We
replicate this analysis for all 57 experimental conditions that purport to
support the group polarization hypothesis. To facilitate this analysis, we
created a web app to manage the data entry, the process of identifying invalid
replications, and managing the outcome data of which replications were
invalidated, and what null model parameters generated the spurious polarization.

## System for identifying invalid replications

The next step is to scale up the analysis we did manually in the example above,
including develop a systematic way to find, and not guess, the set of null model
parameters that

![**Computational system to interactively simulate opinion change, identify
invalid replications, and evaluate experimental
designs.**](Figures/Model/SoftwareMap_Mashup.png){#fig-computational-system
fig-align="center" width="1011"}

# Part 2: *Evaluate* Experimental Designs

In the previous step we identified which, if any, null polarization models
generate spurious group polarization with the expectation model of opinion
measurement. This is enough to invalidate a replication, but it leaves
unanswered questions. First, we would like to know how what the chance was that
a published detection of group polarization was actually a false positive,
assuming that participant opinions were distributed according to the null model
identified in the first part of the analysis. This is equivalent to the false
positive rate of the experimental design, which we denote $\alpha_e$ for
experiment $e$. Second, we would like to estimate the false discovery rate for
our corpus, which is a good proxy for the field as a whole, since we selected
top quality, highly influential work that has been cited thousands of times. The
false discovery rate in this case tells us the probability that a published
detection of group polarization is actually a false detection. (@eq-fdr).

## Generative Model of Spurious Group Polarization

To evaluate an experimental design's ability to properly identify null effects,
we need to simulate participant responses assuming the mere agreement null model
that invalidated the replication, if one was found. We simulate data from a
single group polarization experiment by sampling participants' ordinal opinions
at pre-/post-discussion from the pre- and post-discussion opinion distributions
specified by the null model that invalidated the replication in that experiment,
$\nu_e$. If there are $N_e$ participants in experiment $e$ then our simulated
data sets contained $2N_e$ rows. We recorded $N_e$ in the web app as metadata
along with data about average opinions and opinion shifts for each experimental
condition across the corpus.

It can be helpful to

@@

@@

Each trial $i$ of $N_e$ simulated pre- and post-deliberation observations is
then inspected to see if spurious group polarization occurred in the simulation.
Spurious group polarization is determined by first inferring the latent
parameters of each simulated experimental trial by fitting an ordered probit
model [@Liddell2018]. The procedure infers the latent distribution parameters
that we in fact specified with $\nu$: $\mu_\mathrm{pre}$, $\sigma_\mathrm{pre}$,
$\mu_\mathrm{post}$, and $\sigma_\mathrm{post}$.

The inferred parameters are then used to calculate the effect size for that
trial, Cohen's $d$, defined as\~\cite[p. 331]{Liddell2018} $$
  d = \frac{\mu_\mathrm{post}- \mu_\mathrm{pre}}{\sqrt{\frac{1}{2}(\sigma_\mathrm{post}^2 + \sigma_\mathrm{pre}^2)}}.
$$ {#eq-cohens} It is implied that these values are the inferred values, not the
values from $\nu$, when used in the context of calculating Cohen's $d$. Note
$d=0$ always for the true $\nu$.

@@

To calculate the family-wise error rate across all experimental conditions of a
study, $s$, $$
\alpha_s = \frac{1}{|s|} \sum_{e \in s} \alpha_e,
$$ where $e \in s$ indicates sum over all experimental conditions $e$ presented
in the article associated with study $s$. The mean across all results identified
as plausibly spurious is $$
 \alpha_\mathrm{all} = \frac{1}{|E|} \sum_{e \in E} \alpha_e.
$$

@@

The false discovery rate is the false positive rate, scaled by the rate of true
negative effects and normalized by the frequency of true positive effects\~
(@eq-fdr) and assuming empirically-motivated values for base rate, $b$, and
statistical power, $W$, as well as more pessimistic and optimistic settings for
base rate and power. Using this notation, the false detection rate is

$$
\mathrm{FDR}= \frac{(1 - b)\alpha}{(1 - b)\alpha + bW}.
$$ {#eq-fdr}

# Part 3: *Specify* Significance Threshold for Experimental Design
